Thesis: draft

### 2021-02-16 draft.tex
made some small adaptations

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individuals time line according to what might be relevant to that individual \cite{Perra2019}. It is thus important to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions, while on the other hand we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.

@@

The advantage of working with network models is that they reduce the level of complexity encountered in the real world, so that one can treat these systems in a more practical way. However we do want our models to display properties similar to the ones seen in real systems \cite{Mata2020}. Since many real systems are not static but evolve in time, this means that we don't only need to deal with static networks but also with temporal networks. Temporal networks are networks where the edges are not always active, but instead become active for some periods of time \cite{Holme2012}. In temporal networks time becomes an explicit element in the network representation \cite{Holme2012}. Static networks have been widely studied and are often convenient for their analytic tractability, whereas temporal networks are, in some cases, more realistic \cite{Mata2020}.\\

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}. The basic idea of all opinion dynamic models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are obviously a simplification of real world opinion dynamics. It is however shown that they do display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions A or B and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. \\
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Here popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3} %also give section
a detailed description of the filtering algorithms used in  this thesis will be given.
% add reference to chapter 2
%The rules that determine how an agent updates his or her opinion can include many different aspect, such as the higher or lower influence somebody can have on others, the resistance to change to a new opinion, peer pressure,... The structure of social networks plays an important role in aspects as peer pressure, whereas the algorithmic personalization used on social media platforms may interfere with the information someone is exposed to. Algorithmic personalization is the individualized way in which social media platforms show posts/information to each of its users.

@@

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics.\\
The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals was well know, than the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. \\
However, the use of statistical physics may also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}. Note that this is not only an important factor in social physics, but also in regular physical research and science in general.
% talk about history

@@

As said earlier, these three main characteristics are the same for on- and off-line social networks. Another property that is found in most on- and off-line networks is the appearance of a giant component, this is, most nodes belong to the same connected component \cite{Latapy}. They also often posses a modular/community structure \cite{Ferrara2012}\cite{McGlohon2011}. \\
There are, however, some properties that are different between on-line and off-line networks as well. One of them might be the degree correlations/mixing pattern. It is commonly assumed and accepted that most real-life social networks have an assortative mixing pattern \cite{Hu2009}\cite{Zhang2014}. For a long time, it was thought that this is also valid for on-line social networks. However, recent research has shown that a lot of on-line social networks display disassortative mixture \cite{Hu2009}\cite{Zhang2014}. This discrepancy between real-life social networks and on-line social networks may be explained as follows. In real-life, `ordinary people' often want to be friends with celebrities, however, the celebrities usually prefer to stay within their own circles \cite{Hu2009}\cite{Zhang2014}. The `ordinary people' don't have access to these circles and we are left with the situation where `ordinary people' interact with `ordinary people' and celebrities hang out among other celebrities; hence, an assortative structure. In on-line social networks, `ordinary people' can easily connect with the celebrity and vice-versa, the celebrity wants to show his/her importance/influence by the number of fans \cite{Hu2009}\cite{Zhang2014}. This leads to a disassortative mixing pattern.

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. This chapter focuses on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure. Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (see Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (see Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism may change the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary models and continuous approaches. These will be discussed in the following two sections, Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn people is discussed and, in Section \ref{thesisOp}, a summary of the opinion dynamics model used in this thesis is given.

@@

Binary models are models where each agent can have one of the two opinions A or B (or, equivalently, 0 or 1). A binary model could reflect real-life cases where there exist two competing opinions, such as a vote in an election system with two competing candidates \cite{Nguyen2020}.

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B (or 0 or 1). Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion.\\
The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent has an opinion, then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive such a system to converge to unanimity and does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan (also note that even when the opinions are equally distributed, there may still be a slight imbalance due to the random assignement of an opinion to each node; instead of a perfect 50/50 distribution, one may end up with a eg. 49.8/50.2 distribution; in this sense there is still some kind of imbalance and a notion of `minority' and `majority' opinion)\cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}.
% https://arxiv.org/pdf/0710.3256.pdf Contains a mean field mathimatical description, interesting?

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fractions of neighbors of user $i$ with opinion A and B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors has opinion B), user $i$ will updates his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that one can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{Modelling opinion dynamics in the age of algorithmic personalization} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from its friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time line; the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the users and then selects the first $Q$ elements
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC method is mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.

>>>>

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individuals time line according to what might be relevant to that individual \cite{Perra2019}. It is thus important to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions, while on the other hand we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.
% refer to relevant sections in introduction?

@@

The advantage of working with network models is that they reduce the level of complexity encountered in the real world, so that one can treat these systems in a more practical way. However, we do want our models to display properties similar to the ones seen in real systems \cite{Mata2020}. Since many real systems are not static but evolve in time, this means that we don't only need to deal with static networks but also with temporal networks. Temporal networks are networks where the edges are not always active, but instead become active for some periods of time \cite{Holme2012}. In temporal networks time becomes an explicit element in the network representation \cite{Holme2012}. Static networks have been widely studied and are often convenient for their analytic tractability, whereas temporal networks are, in some cases, more realistic \cite{Mata2020}.\\

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
The basic idea of all opinion dynamic models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are obviously a simplification of real world opinion dynamics. It is however shown that they do display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions A or B and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn people will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Here popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in  this thesis will be given.
% add reference to chapter 2
%The rules that determine how an agent updates his or her opinion can include many different aspect, such as the higher or lower influence somebody can have on others, the resistance to change to a new opinion, peer pressure,... The structure of social networks plays an important role in aspects as peer pressure, whereas the algorithmic personalization used on social media platforms may interfere with the information someone is exposed to. Algorithmic personalization is the individualized way in which social media platforms show posts/information to each of its users.

@@

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics.\\
The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, then the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. \\
However, the use of statistical physics may also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}. Note that this is not only an important factor in social physics, but also in regular physical research and science in general.
% talk about history

@@

As said earlier, these three main characteristics are valid for on- and off-line social networks. Another property that is found in most on- and off-line networks is the appearance of a giant component, this is, most nodes belong to the same connected component \cite{Latapy}. They also often posses a modular/community structure \cite{Ferrara2012}\cite{McGlohon2011}. \\
There are, however, some properties that are different between on-line and off-line networks as well. One of them might be the degree correlations/mixing pattern. It is commonly assumed and accepted that most real-life social networks have an assortative mixing pattern (this distinguishes them from other real-world networks, eg. biological networks \cite{F.Costa2007}, which often have a disassortative pattern) \cite{Hu2009}\cite{Zhang2014}. For a long time, it was thought that this is also valid for on-line social networks. However, recent research has shown that a lot of on-line social networks display disassortative mixture \cite{Hu2009}\cite{Zhang2014}. This discrepancy between real-life social networks and on-line social networks may be explained as follows. In real-life, `ordinary people' often want to be friends with celebrities, however, the celebrities usually prefer to stay within their own circles \cite{Hu2009}\cite{Zhang2014}. The `ordinary people' don't have access to these circles and we are left with the situation where `ordinary people' interact with `ordinary people' and celebrities hang out among other celebrities; hence, an assortative structure. In on-line social networks, `ordinary people' can easily connect with the celebrity and vice-versa, the celebrity wants to show his/her importance/influence by the number of fans \cite{Hu2009}\cite{Zhang2014}. This leads to a disassortative mixing pattern.

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. This chapter focuses on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure. Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (see Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (see Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism may change the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary/discrete models and continuous approaches. These will be discussed in the following two sections, Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn people is discussed and, in Section \ref{thesisOp}, a summary of the opinion dynamics model used in this thesis is given.

@@

Binary models are models where each agent can have one of the two opinions A or B (or, equivalently, 0 or 1). A binary model could reflect real-life cases where there exist two competing opinions, such as a vote in an election system with two competing candidates or choosing between Windows and Linux \cite{Nguyen2020}.

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B (or 0 or 1). Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion.\\
The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent has an opinion, then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive such a system to converge to unanimity; and, does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? \\
Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan (also note that even when the opinions are equally distributed, there may still be a slight imbalance due to the random assignement of an opinion to each node; instead of a perfect 50/50 distribution, one may end up with a eg. 49.8/50.2 distribution; in this sense there is still some kind of notion of `minority' and `majority' opinion) \cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}.
% https://arxiv.org/pdf/0710.3256.pdf Contains a mean field mathimatical description, interesting?

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fractions of neighbors of user $i$ with opinion A and B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors has opinion B), user $i$ will updates his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that human beings can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{Modelling opinion dynamics in the age of algorithmic personalization} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time line; the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $Q$ elements
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC method is mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.
